Learning Discrete-valued Bayesian Networks from Mixed Data

### 2015-11-17 BN_discretizer_paper.tex
remerging Mykel's comments

\newcommand{\todo}[1]{\textcolor{magenta}{#1}}
\newcommand{\tim}[1]{\textit{\textcolor{blue}{#1}}}

@@

\section{Introduction}
\label{intro}
Bayesian networks \citep{Pearl_1988, PGM_2009} are an increasingly popular method for modeling uncertainty and causality in science and engineering. They provide an efficient factorization of the joint probability distribution over a set of random variables. Bayesian networks first emerged from artificial intelligence research and have been applied to a wide variety of problems, ranging from decision-making systems \citep{DMU_2015} to medical diagnoses \citep{Lustgarten_2011}. In most cases, we assume that all random variables in Bayesian networks are discrete, since many Bayesian network learning algorithms are unable to deal with continuous variables efficiently. However, many applications
require the use of continuous variables, such as position and velocity in dynamic systems.

There are two common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters.
One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}.
For other parametric distributions \citep{Ihler_2009} and non-linear functions~\todo{[cite]}, these redesigned algorithms are computationally expensive and do not perform well.

@@

\subsection{MDL Principle Discretization}
The MDL principle was first proposed by \cite{MDL_1978}, and states that the best model for a dataset is one which minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and optimal the positions of their discretization boundaries.
Their approach selects a discretization policy that minimizes the sum of the description lengths of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.
In the single continuous variable case, the objective function is

@@

% The discretization technique proposed by \citet{Friedman_1996} is the most well-known of these methods. It is based on the minimum description length principle (MDL): the optimal discretization policy minimizes the description length of the Bayesian network and discretization policy. The MDL algorithm has a $\bigo\paren{N^3 + \paren{n_c {v_\text{max}}^{(n_c^p)_\text{max}} + {v_\text{max}}^{n_p}} \cdot N^2}$ runtime on Bayesian networks with a single continuous variable and multiple discrete variables, where $N$ is the number of data samples from the joint distribution, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_\text{max}$ is the largest cardinality over all variables in the Markov blanket, and  ${(n_c^p)_\text{max}}$ is the largest number of parent variables of the continuous variable's children. The MDL method iterates over all continuous variables until convergence in the case of multiple continuous variables, converging in a few cycles according to our tests of real-world data. Unfortunately, MDL tends to produce discrete variables with low cardinality~\todo{[cite]}.

@@

A Bayesian network $B$ is defined by a pair $(G,\Theta)$, where $G = (X,E)$ is a directed acyclic graph whose nodes correspond to a set of random variables $X = \brock{X_1, X_2, \cdots, X_n}$, and whose edges $E$ represent probabilistic dependencies among nodes. The graph structure $G$ encodes the Markov property: each node $X_i$ is independent of its non-descendants given its parents $\Pi_{X_i}$ in $G$. The conditional relations are parameterized by the elements of $\Theta$, which take the form $\theta_{x_i \mid \Pi_{x_i}} = P(x_i \mid \Pi_{x_i})$ for each possible value $x_i$ of $X_i$, and $\Pi_{x_i}$ of $\Pi_{X_i}$. The joint distribution over $X$ factors according to the Markov property:
\begin{equation}
P_B (X_1 , \cdots, X_n) = \prod_{i=1}^{n} P_B (X_i \mid \Pi_{X_i}) = \prod_{i=1}^{n} \theta_{x_i \mid \Pi_{x_i}}\text{.}
\end{equation}

Forward sampling is used to draw samples from the joint distribution represented by a Bayesian network.
Variables are sampled in topological order such that parental instantiations are always known~\citep[see][chap.~22]{algo_2009}.
Furthermore, the parent $\rightarrow$ child sampling order can be reversed if the marginal probability of the child variable is known.
By Bayes' rule,

\begin{equation}
P(X_i \mid \Pi_{X_i}) \cdot \prod_{j = 1}^{ Pa( X_i)} P( \{ \Pi_{X_i} \}_j) = P(X_i) \cdot P(\Pi_{X_i} \mid X_i),
\end{equation}

\noindent
where $Pa(X_i)$ is the number of parents of $X_i$ and $\brock{\Pi_{X_i}}_j$ is the $j$th parent of $X_i$.
% One can first sample $X_i$ by $P(X_i)$, then sample all the parent variables simultaneously by $P(\Pi_{X_i} \mid X_i)$. That is to say, $X_i$ becomes the starting point of sampling.

@@

Every midpoint between two nonequal consecutive values in $D_X$ is a candidate discretization edge.
Let ${D'_X} = \brock{x'_1,x'_2,\ldots,x'_M}$ be the set of unique values of $D_X$ sorted in ascending order, where $M$ is the number unique values in $D_X$.
Recall that $D_X$ is also sorted.
Let $b = \brock{b_0,b_1,b_2,\ldots,b_M}$ be an increasing sequence of integers marking the repeated value intervals in $D_X$ such that $b_0 = 0$ and $x_{b_{i-1} + 1} = x_{b_{i-1} + 2} = \ldots = x_{b_i} = x'_i$.
By this definition, the allowable discretization positions are $d_i = \paren{x_{b_{i}} + x_{{b_i}+1}}/2$ for all $i = 1,2,\ldots,M-1$.

@@

The evaluation of $h(u,v)$ for all $u \leq v$ is summarized in Algorithm \ref{alg:h} in the appendix.
The calculation has a $O(n_c  {v'_\text{max}} \cdot N^2 + {v'_\text{max}}^{n_p} \cdot N^2)$ runtime, where $n_c$ and $n_p$ are the numbers of child and parent variables respectively, and $v'_\text{max}$ is the largest cardinality of variables that directly connect to $X$.

Now we are able to solve the optimization problem over Equation \ref{eq:opt_prob}.
The dynamic programming procedure is shown in Algorithm \ref{alg:disc_one}.
It takes two inputs: $D$, the joint samples over all variables sorted in ascending order according to $D_X$, and $G$, the network structure.
The runtime of Algorithm \ref{alg:disc_one} is also $O(n_c  {v'_\text{max}} \cdot N^2 + {v'_\text{max}}^{n_p} \cdot N^2)$, because the runtime of the dynamic programming procedure is less than the runtime for computing $h(u,v)$.
Algorithm \ref{alg:disc_one} is guaranteed to be optimal.
For faster methods with suboptimal results please refer to \citep{Boulle_2006}.

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm \ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $M$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\State
\For {$v = 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u = 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $M$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\section{Extension to Multiple Continuous Variables}
\label{sec:multi_var}

@@

\begin{algorithm}
  \caption{ Learning a discrete-valued Bayesian network}
  \label{alg:structure_learn}
  \begin{algorithmic}[5]
  \Function{Learn\_DBN}{$D''$, $C$, $order$, $u_{parent}$, $C$, $u_{cycle}$}
    \State $n \leftarrow$ the number of variables in the Bayesian network
    \State $k \leftarrow median\{ \|v\|, v\notin C\}$
    \State $M[i] \leftarrow$ the discretization policy for the $i$th variable
    \State $D^*_i \leftarrow M[i] (D_i)$, list of samples $D_i$ discretized according to $M[i]$
    \State $D^* \leftarrow $ the discretized data for all variables, where $D^*_i \leftarrow D''_i$ if $i \notin C$
    \State $G \leftarrow$ the initial edgeless graph structure
    \For {$i = 1$ to $n$}
      \If {$i \in C$}
        \State $M[i] \leftarrow$  equal-width discretization with $k$ intervals
        \State $D^*_i \leftarrow  M[i] (D''_i)$
      \EndIf
    \EndFor
    \For {$i =1$ to $n$}
      \State $P_{old} \leftarrow f(X^*_i,\Pi_{X^*_i})$: Equation \ref{eq:f_x_p}
      \State OKToProceed $\leftarrow$ \textbf{true}
      \While {OKToProceed \textbf{and} $\|\Pi_{X^*_i}\| < u_{parent}$}
        \State $Y \leftarrow$ an element from the set $order[1:i] \backslash \Pi_X$
        \State $P_{new} \leftarrow f(X^*_i,\Pi_{X^*_i} \cup Y)$
        \If {$P_{new} > P_{old}$}
          \State $P_{old} \leftarrow P_{new}$
          \State $\Pi_{X_i} \leftarrow \Pi_{X_i} \cup Y $
          \State $M \leftarrow$ {DISCRETIZE}({$D''$, $G$, $C$, $u_{cycle}$}): Algorithm \ref{alg:disc_two}
          \State $D* = M(D'')$
        \Else
          \State OKToProceed $\leftarrow$ \textbf{false}
        \EndIf
      \EndWhile
    \EndFor
    \State \Return $G$, $M$
  \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\subsubsection{Experiment 1 on the Auto MPG dataset}
\label{subsubsec:auto_exp1}

@@

\subsubsection{Experiment 2 on the Auto MPG dataset}
\label{subsubsec:auto_exp2}

@@

\subsubsection{Experiment 1 on the Wine dataset}
\label{subsubsec:wine_exp1}

The Bayesian network in Figure \ref{fig:wine_graph_1} is obtained by equal-width discretizing each continuous variable into 3 intervals and running K2 1000 times. Table \ref{table:wine_disc_table_1} shows a comparison between the proposed method and the MDL principle discretization method. The proposed method is still better than MDL principle discretization in terms of likelihood, although the latter returns non-trivial result in this experiment. Note that the proposed method with the approximation has more intervals after discretization, and it has a slightly better result in terms of likelihood. However, the improvement is marginally small, which indicates that the proposed method without the approximation captures good enough discretization result even if the numbers of intervals after discretization are smaller. In addition, as shown in next dataset, the approximation might cut continuous variables into too many intervals. Therefore, the proposed method without the approximation should be still considered a better approach.

\begin{figure}[ht]
\centering
    \begin{tabular}{cc}
      \input{graph_wine_1}
    \end{tabular}
  \caption{K2 structure learning with prediscretization on the Wine dataset}
  \label{fig:wine_graph_1}
\end{figure}

\begin{table}
\centering
\scalebox{1.0}{
\input{table_wine_1.tex}
}
\caption{Discretization result of Wine dataset based on the graph Fig. \ref{fig:wine_graph_1}}
\label{table:wine_disc_table_1}
\end{table}

@@

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_wine_exp1_distr}
    \end{tabular}
  \caption{Comparison of the discretization results on variable $A$ and $C$ obtained by the proposed method and the MDL method}
  \label{fig:wine_exp1_distr}
\end{figure}

\subsubsection{Experiment 2 on the Wine dataset}
\label{subsubsec:wine_exp2}

Figure \ref{fig:wine_graph_2} is the learned discrete-valued Bayesian network from the Wine dataset, and it is obtained by running Algorithm \ref{alg:structure_learn} 50 times. The integer numbers next to variables are cardinality numbers of variables. By comparing Figure \ref{fig:wine_graph_2} with \ref{fig:wine_graph_1}, it shows that the the Bayesian network obtained by \ref{alg:structure_learn} has more edges than the network obtained by prediscretization and the K2 algorithm. Furthermore, the cardinalitis of discretized variables in Figure \ref{fig:wine_graph_2} are smaller than the result in \ref{table:wine_disc_table_1}.

@@

\begin{figure}[ht]
    \begin{tabular}{cc}
    \centering
    \scalebox{0.7}{
      \input{graph_wine_2}}
    \end{tabular}
  \caption{The learned discrete-valued Bayesian network from the Wine dataset}
  \label{fig:wine_graph_2}
\end{figure}

\begin{figure}[ht]
    \begin{tabular}{cc}
      \input{graph_wine_exp2_distr_5_11}
    \end{tabular}
  \caption{The discretization result on variable $E$ and $K$ of the learned network in Figure \ref{fig:wine_graph_2}}
  \label{fig:wine_exp2_distr}
\end{figure}

@@

\subsubsection{Experiment 1 on the Housing dataset}
\label{subsubsec:housing_exp1}

The Bayesian network structure in Figure \ref{fig:housing_graph_1} is obtained by prediscretizing each continuous variable into five uniform-width intervals, then running the K2 algorithm one thousands time and choosing the network with the highest K2 score. Table \ref{table:housing_disc_table_1} shows the numbers of intervals after discretization on each continuous variables. MDL method does not discretize most variables. The proposed method with the approximation discretizes variables into too many intervals. For example, it discretized $J$ into 58 intervals. The proposed method without the approximation has a good control of number of intervals.
\begin{figure}[ht]
\centering
    \begin{tabular}{cc}
      \input{graph_housing_1}
    \end{tabular}
  \caption{K2 structure learning with prediscretization on Housing dataset}
  \label{fig:housing_graph_1}
\end{figure}

\begin{table}
\centering
\scalebox{1.0}{
\input{table_housing_1.tex}
}
\caption{Discretization result of Housing dataset based on the graph Fig. \ref{fig:housing_graph_1}}
\label{table:housing_disc_table_1}
\end{table}

>>>>

\newcommand{\todo}[1]{\textcolor{magenta}{#1}}
\newcommand{\yichun}[1]{\textit{\textcolor{green}{#1}}}
\newcommand{\tim}[1]{\textit{\textcolor{blue}{#1}}}
\newcommand{\mykel}[1]{\textit{\textcolor{cyan}{#1}}}

@@

\section{Introduction}

\begin{table}[H]
  \centering
  \begin{tabular}{@{}ccl@{}}
    \toprule
    Current & Mykel & Description \\
    \midrule
    $G$ & $G$ & BN graph structure \\
    $P$ & $P$ & discrete probability \\
    $p$ & $p$ & probability density \\
    $D$ & $D$ & available data \\
    $|D|$ & $n$ & number of variables in the BN \\
    $D_{\bm{Y}}$ & & subset of data associated with $\bm{Y}$ \\
    $X$ & $X_{1:n}$ & variables in BN (discrete) \\
    $x_{1:n}$ & $x_{1:n}$ & a specific instance of vars in BN \\
    $x'_{1:n}$ &  & unique instances of var in BN \\
        & $r_i$ & number of instantiations of $i$th var \\
    $J_p$ & $q_i$ & number of instantiations of parents of $i$th var \\
        &       & $ = 1$ if no parents \\
    $\brock{\Pi_{X_i}}_j$ & $\pi_{ij}$ & $j$th instantiation of $i$th var \\
        & $\theta_{ijk}$ & probability of $i$th var $= k$ given $j$th parental inst. \\
    $\Theta$ & $\bm{\theta}$ & set of all (non-independent) probability parameters \\
    $n_{i,j}^{(k)}$ & $m_{ijk}$ & observation counts \\
    $n_{i,j,l}^{(k)}$ & $m_{ijk}$ & observation counts with $l$th instance of spouse\\
    $\alpha_{ijk}$ & $\alpha_{ijk}$ & Dirichlet prior counts \\
    $N_{ij}$ & $\alpha_{ij0}$ & sum over Dirichlet prior counts for $k$ \\
        & $m_{ij0}$ & sum over observations counts for $k$ \\
        & $\Gamma$ & the Gamma function \\
    $N$ &          & is the number of samples in dataset \\
    $N_i$ &          & counts for $i$th instance of a specific var \\
    $\Pi_X$ & $\text{pa}_{x_i}$ & parents of a variable or instance \\
    $n_C$ & & number of children of a variable \\
    $H(~)$ & & a function \\
    $I(~,~)$ & & mutual information \\
    $X_i$ & & $i$th variable in BN, may be cont or disc \\
    $X_i^*$ & & discrete version of cont var \\
    $E$ & & set of edges in BN \\
    $\text{Pa}\paren{X_i}$, $n_p$ & & number of parents of $X_i$ \\
    $f\paren{X_i, \Pi_{X_i}}$ & $f(G)$ & score function for K2 \\
    $M_C$ & & discretization policy for $C$ \\
    $C$ & & a continuous variable \\
    $t_i$ & & the $i$th bin edge for a specific var \\
    $d_i$ & & allowable discretization positions (between unique vals) \\
    $k$ & & number of intervals for $M_C$ \\
    $C_{1:N}$ & & sorted samples for $C$ \\
    $n_i$ & & $n_1 = s_1$ and $n_i = s_{i} - s_{i-1}$ \\
    $s_i$ & & $\sum_{j=1}^k n_j$ and $s_k = N$, $s_0 = 0$ \\
    $C_i$ & & $i$th child of a specific var \\
    $\boldsymbol{S_i}$ & & set of spouses associated with $i$th child of specific var \\
    $L$ & & largest cardinality among discrete variables in BN \\
    $I_q$ & & an interval from $u \rightarrow v$ \\
    $v'_\text{max}$ & & largest cardinality of variables that directly connect to $X$ \\
    $S[]$ & &  optimal objective value of subproblems with instances from $1$ to $u$ \\
    \bottomrule
  \end{tabular}
\end{table}

Bayesian networks \citep{Pearl_1988, PGM_2009} are an increasingly popular method for modeling uncertainty and causality in science and engineering.
They provide an efficient factorization of the joint probability distribution over a set of random variables.
Bayesian networks first emerged from artificial intelligence research and have been applied to a wide variety of problems, ranging from decision-making systems \citep{DMU_2015} \mykel{better to use a primary source} to medical diagnoses \citep{Lustgarten_2011}.
In most cases, we assume that all random variables in Bayesian networks are discrete, since many Bayesian network learning algorithms are unable to deal with continuous variables efficiently.
However, many applications   +Bayesian networks \citep{Pearl_1988, PGM_2009} are an increasingly popular method for modeling uncertainty and causality in science and engineering.
They provide an efficient factorization of the joint probability distribution over a set of random variables.
Bayesian networks first emerged from artificial intelligence research and have been applied to a wide variety of problems, ranging from decision-making systems \citep{DMU_2015} to medical diagnoses \citep{Lustgarten_2011}.
In most cases, we assume that all random variables in Bayesian networks are discrete, since many Bayesian network learning algorithms are unable to deal with continuous variables efficiently.
However, many applications require the use of continuous variables, such as position and velocity in dynamic systems \mykel{cite Kochenderfer JGCD article on encounter modeling}.

There are two common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning algorithms based on their parameters.
One successful example is belief propagation in Gaussian graphical models \citep{Weiss_2011}.
For other parametric distributions \citep{Ihler_2009} and non-linear functions~\todo{[cite]}, these redesigned algorithms are computationally expensive and do not perform well.
\mykel{$\bigstar$ We need to discuss this; talk about algorithms for inference and structure learning, expectation propagation, Gibbs, BVGS}

@@

\subsection{MDL Principle Discretization}
The MDL principle was first proposed by \cite{MDL_1978} and states that the best model for a dataset is one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization boundaries.
Their approach selects a discretization policy that minimizes the sum of the description lengths of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.
For a Bayesian network with a single continuous variable $X$, the objective function is

@@

% The discretization technique proposed by \citet{Friedman_1996} is the most well-known of these methods. It is based on the minimum description length principle (MDL): the optimal discretization policy minimizes the description length of the Bayesian network and discretization policy. The MDL algorithm has an $\bigo\paren{N^3 + \paren{n_c {v_\text{max}}^{(n_c^p)_\text{max}} + {v_\text{max}}^{n_p}} \cdot N^2}$ runtime on Bayesian networks with a single continuous variable and multiple discrete variables, where $N$ is the number of data samples from the joint distribution, $n_p$ and $n_c$ are the numbers of parent and children variables for the continuous variable, $v_\text{max}$ is the largest cardinality over all variables in the Markov blanket, and  ${(n_c^p)_\text{max}}$ is the largest number of parent variables of the continuous variable's children. The MDL method iterates over all continuous variables until convergence in the case of multiple continuous variables, converging in a few cycles according to our tests of real-world data. Unfortunately, MDL tends to produce discrete variables with low cardinality~\todo{[cite]}
\mykel{don't like low-cardinality. Maybe 'intervals' or something instead}.

@@

\mykel{$\bigstar$ This all needs to be explained better. Too many variables are being defined at once.}

A Bayesian network $B$ is defined by a pair $(G,\Theta)$, where $G = (X,E)$ is a directed acyclic graph whose nodes correspond to a set of random variables $X = \brock{X_1, X_2, \ldots, X_n}$, and whose edges $E$ represent probabilistic dependencies among nodes.
The graph structure $G$ encodes the Markov property: each node $X_i$ is independent of its non-descendants given its parents $\Pi_{X_i}$ in $G$.
The conditional relations are parameterized by the elements of $\Theta$, which take the form $\theta_{x_i \mid \Pi_{x_i}} = P(x_i \mid \Pi_{x_i})$ for each possible value $x_i$ of $X_i$, and $\Pi_{x_i}$ of $\Pi_{X_i}$. The joint distribution over $X$ factors according to the Markov property:

\begin{equation}
P_B (X_1 , \ldots, X_n) = \prod_{i=1}^{n} P_B (X_i \mid \Pi_{X_i}) = \prod_{i=1}^{n} \theta_{x_i \mid \Pi_{x_i}}\text{.}
\end{equation}

Forward sampling is used to draw samples from the joint distribution represented by a Bayesian network.
Variables are sampled in topological order such that parental instantiations are always known~\citep[see][chap.~22]{algo_2009}.
\mykel{I think this ref is just for the topological sort; for sampling, cite DMU or PGM}
Furthermore, the parent $\rightarrow$ child sampling order can be reversed if the marginal probability of the child variable is known.
By Bayes' rule,

\begin{equation}
P(X_i \mid \Pi_{X_i}) \cdot \prod_{j = 1}^{ Pa( X_i)} P( \brock{\Pi_{X_i} }_j) = P(X_i) \cdot P(\Pi_{X_i} \mid X_i)\text{,}
\end{equation}

\noindent
where $Pa(X_i)$ is the number of parents of $X_i$ and $\brock{\Pi_{X_i}}_j$ is the $j$th parent of $X_i$.
\mykel{this may be messed up. Maybe use DMU notation?}
% One can first sample $X_i$ by $P(X_i)$, then sample all the parent variables simultaneously by $P(\Pi_{X_i} \mid X_i)$. That is to say, $X_i$ becomes the starting point of sampling.

@@

Every midpoint between two nonequal consecutive values in $D_X$ is a candidate discretization edge.
Let ${D'_X} = \brock{x'_1,x'_2,\ldots,x'_M}$ be the set of unique values of $D_X$ sorted in ascending order, where $M$ is the number unique values in $D_X$.
Recall that $D_X$ is also sorted.
Let $b = \brock{b_0,b_1,b_2,\ldots,b_M}$ be an increasing sequence of integers marking the repeated value intervals in $D_X$ such that $b_0 = 0$ and $x_{b_{i-1} + 1} = x_{b_{i-1} + 2} = \ldots = x_{b_i} = x'_i$.
By this definition, the allowable discretization positions are $d_i = \paren{x_{b_{i}} + x_{{b_i}+1}}/2$ for all \todo{$i = 1,2,\ldots,M-1$}.

@@

The evaluation of $h(u,v)$ for all $u \leq v$ is summarized in Algorithm \ref{alg:h} in the appendix.
The calculation has a $O(n_c  {v'_\text{max}} \cdot N^2 + {v'_\text{max}}^{n_p} \cdot N^2)$ runtime, where $n_c$ and $n_p$ are the numbers of child and parent variables respectively, and $v'_\text{max}$ is the largest cardinality of variables that directly connect to $X$. \tim{so parents and children, or parents, children and spouses?}

The optimization problem over Equation~\ref{eq:opt_prob} can now be solved.
The dynamic programming procedure is shown in Algorithm~\ref{alg:disc_one}.
It takes two inputs: $D$, the joint samples over all variables sorted in ascending order according to $D_X$, and $G$, the network structure.
The runtime of Algorithm~\ref{alg:disc_one} is also $O(n_c  {v'_\text{max}} \cdot N^2 + {v'_\text{max}}^{n_p} \cdot N^2)$, because the runtime of the dynamic programming procedure is less than the runtime for computing $h(u,v)$.
Algorithm~\ref{alg:disc_one} is guaranteed to be optimal.
For faster methods with suboptimal results please refer to \citep{Boulle_2006}.

\begin{algorithm}
\caption{Discretization of one continuous variable in a Bayesian network}
\label{alg:disc_one}
\begin{algorithmic}[5]
\Function{Discretize}{$D$, $G$}
\State
\State $N \leftarrow$ the number of instances
\State $N' \leftarrow$ the number of unique values in $D_X$
\State $H \leftarrow$ an $N \times N$ matrix such that $H[u,v] = h(u,v)$ per Algorithm~\ref{alg:h} in the appendix
\State $b \leftarrow$ the increasing sequence of $M$ integers as previously defined
\State $L \leftarrow$ the largest cardinality over all discrete variables in the Markov blanket of $X$
\State $S[u] \leftarrow$ the optimal objective value of subproblems with instances from $1$ to $u$
\State $M[u] \leftarrow$ the optimal discretization for subproblems with instances from $1$ to $u$
\State $W[u]  \leftarrow - \log\brock{1 - {\exp\paren{- L \cdot{ {{x_{b(u)+1} - x_{b(u)}}\over{x_N - x_1}}}}}}$ for $u = 1,2, \ldots,N'-1$ and $L(N') \leftarrow 0$
\State
\For {$v = 1$ to $N'$}
\If {$v = 1$}
\State $S[v] \leftarrow g \left(1,b[v] \right) + L[v]$
\State $M[v] \leftarrow \{ ({x_{b[v]} + x_{b[v]+1}}) / 2\}$
\Else
\State $s \leftarrow \infty$ and $boundary \leftarrow \infty$
\For {$u = 1$ to $v$}
\State $s' \leftarrow S[u] + g \left( b[u]+1,b[v] \right) +  {L \cdot {{x_{b[v]} - x_{b[u] + 1}}\over{x_N - x_1}}} + W[v]$
\If {$s' < s$}
\State $s \leftarrow s'$
\State $boundary \leftarrow ({x_{b[u]} + x_{b[u]+1}}) / 2$
\EndIf
\EndFor
\State $S[v] \leftarrow s$
\State $M[v] \leftarrow M[u] \cup \{ boundary\}$
\EndIf
\EndFor
 \State \Return $M$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\section{Multiple Variable Discretization}
\label{sec:multi_var}

@@

\begin{algorithm}
  \caption{ Learning a discrete-valued Bayesian network}
  \label{alg:structure_learn}
  \begin{algorithmic}[5]
  \Function{Learn\_DBN}{$D''$, $C$, $order$, $u_{parent}$, $C$, $u_{cycle}$}
    \State $n \leftarrow$ the number of variables in the Bayesian network
    \State $k \leftarrow median\{ \|v\|, v\notin C\}$
    \State $M[i] \leftarrow$ the discretization policy for the $i$th variable
    \State $D^*_i \leftarrow M[i] (D_i)$, list of samples $D_i$ discretized according to $M[i]$
    \State $D^* \leftarrow $ the discretized data for all variables, where $D^*_i \leftarrow D''_i$ if $i \notin C$
    \State $G \leftarrow$ the initial edgeless graph structure
    \For {$i = 1$ to $n$}
      \If {$i \in C$}
        \State $M[i] \leftarrow$  equal-width discretization with $k$ intervals
        \State $D^*_i \leftarrow  M[i] (D''_i)$
      \EndIf
    \EndFor
    \For {$i =1$ to $n$}
      \State $P_{old} \leftarrow f(X^*_i,\Pi_{X^*_i})$: Equation~\ref{eq:f_x_p}
      \State OKToProceed $\leftarrow$ \textbf{true}
      \While {OKToProceed \textbf{and} $\|\Pi_{X^*_i}\| < u_{parent}$}
        \State $Y \leftarrow$ an element from the set $order[1:i] \backslash \Pi_X$
        \State $P_{new} \leftarrow f(X^*_i,\Pi_{X^*_i} \cup Y)$
        \If {$P_{new} > P_{old}$}
          \State $P_{old} \leftarrow P_{new}$
          \State $\Pi_{X_i} \leftarrow \Pi_{X_i} \cup Y $
          \State $M \leftarrow$ {DISCRETIZE}({$D''$, $G$, $C$, $u_{cycle}$}): Algorithm~\ref{alg:disc_two}
          \State $D* = M(D'')$
        \Else
          \State OKToProceed $\leftarrow$ \textbf{false}
        \EndIf
      \EndWhile
    \EndFor
    \State \Return $G$, $M$
  \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\subsubsection{Discretization with fixed Structure}
\label{subsubsec:auto_exp1}

@@

\subsubsection{Discretization while Learning Structure}
\label{subsubsec:auto_exp2}

@@

\subsubsection{Discretization with fixed Structure}
\label{subsubsec:wine_exp1}

The Bayesian network in Figure \ref{fig:wine_graph_1} is obtained by equal-width discretizing each continuous variable into 3 intervals and running K2 1000 times. Table \ref{table:wine_disc_table_1} shows a comparison between the proposed method and the MDL principle discretization method.
The proposed method is still better than MDL principle discretization in terms of likelihood, although the latter returns non-trivial result in this experiment.
\todo{Note that the proposed method with the approximation has more intervals after discretization, and it has a slightly better result in terms of likelihood. However, the improvement is marginally small, which indicates that the proposed method without the approximation captures good enough discretization result even if the numbers of intervals after discretization are smaller.}
In addition, as shown in next dataset, the approximation might cut continuous variables into too many intervals. Therefore, the proposed method without the approximation should be still considered a better approach.

\begin{figure}[ht]
  \centering
    \begin{tabular}{cc}
      \input{graph_wine_1}
    \end{tabular}
  \caption{\tim{Structure from prediscretized K2 structure learning to compare the methods}}
  \label{fig:wine_graph_1}
\end{figure}

\begin{table}
  \centering
  \scalebox{1.0}{
  \input{table_wine_1.tex}
  }
  \caption{Discretization result of Wine dataset based on the graph Fig. \ref{fig:wine_graph_1}
  \todo{MDL has bins, heuristic is slightly worse in logl but still good. MDL keeps all of its CPDs small, no neighboring vars both have more than 1 interval.}
  \tim{Should we scale the logl? What is the best way to do that?}
  }
  \label{table:wine_disc_table_1}
\end{table}

@@

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph_wine_exp1_distr}
  \end{tabular}
  \caption{Comparison of the discretization results on variable $A$ and $C$ obtained by the proposed method and the MDL method}
  \label{fig:wine_exp1_distr}
\end{figure}

\subsubsection{Discretization while Learning Structure}
\label{subsubsec:wine_exp2}

Figure \ref{fig:wine_graph_2} is the learned discrete-valued Bayesian network from the Wine dataset, and it is obtained by running Algorithm \ref{alg:structure_learn} fifty times. The integer numbers next to variables are cardinality numbers of variables. By comparing Figure \ref{fig:wine_graph_2} with \ref{fig:wine_graph_1}, it shows that the the Bayesian network obtained by Algorithm \ref{alg:structure_learn} has more edges than the network obtained by prediscretization and the K2 algorithm. Furthermore, the cardinalitis of discretized variables in Figure \ref{fig:wine_graph_2} are smaller than the result in \ref{table:wine_disc_table_1}.

@@

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
  \centering
  \scalebox{0.7}{
    \input{graph_wine_2}}
  \end{tabular}
  \caption{The learned discrete-valued Bayesian network from the Wine dataset \todo{using optimal Bayesian method}}
  \label{fig:wine_graph_2}
\end{figure}

\begin{figure}[ht]
  \centering
  \begin{tabular}{cc}
    \input{graph_wine_exp2_distr_5_11}
  \end{tabular}
  \caption{The discretization result on variable $E$ and $K$ of the learned network in Figure \ref{fig:wine_graph_2}
  \todo{Note unchanged discretization edge.}
  }
  \label{fig:wine_exp2_distr}
\end{figure}

@@

\subsubsection{Discretization with Fixed Structure}
\label{subsubsec:housing_exp1}

The Bayesian network structure in Figure \ref{fig:housing_graph_1} is obtained by prediscretizing each continuous variable into five uniform-width intervals, then running the K2 algorithm one thousands time and choosing the network with the highest K2 score. Table \ref{table:housing_disc_table_1} shows the numbers of intervals after discretization on each continuous variables. MDL method does not discretize most variables. The proposed method with the approximation discretizes variables into too many intervals. For example, it discretized $J$ into 58 intervals. The proposed method without the approximation has a good control of number of intervals.
\begin{figure}[ht]
  \centering
  % \begin{tabular}{cc}
    \input{graph_housing_1}
  % \end{tabular}
  \caption{\todo{Structure from prediscretized K2 structure learning to compare the methods for Housing dataset}}
  \label{fig:housing_graph_1}
\end{figure}

\begin{table}
  \centering
  \scalebox{1.0}{
  \input{table_housing_1.tex}
  }
  \caption{Discretization result of Housing dataset based on the graph Figure~\ref{fig:housing_graph_1} \tim{Fig. vs. Figure}}
  \label{table:housing_disc_table_1}
\end{table}
